### 2022-12-24 tex/semirings.tex
[ edit ] another draft of chapters 5 and 6

The rest of this chapter proceeds as follows.
In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared previously in my work with Atkey~\cite{WA21}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
The idea of \name{} is to augment the simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
Given this new calculus $\name$, the first goal is to apply the techniques of
\cref{sec:simple} to it, yielding a simultaneous substitution operation.
To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
the typing rules of \name{} to not mention contexts explicitly, as was the style
in \cref{sec:simple}.
This new notation --- the \emph{bunched connectives} --- is versatile at
defining simply typed usage-aware syntaxes, and I give further non-$\name$
examples in \cref{sec:rec}.
The application of the techniques of \cref{sec:simple} is left to
\cref{sec:ren-sub-lr}.

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{Additions to and variations of \name{}}\label{sec:variant}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

@@

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop its syntactic metatheory up to and including
substitution.
I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
significant changes to the underlying notion of \emph{environment} before doing
so.
I give and informally motivate these changes to environments in
\cref{sec:lrkits}, and prove some properties of the new definition in
\cref{sec:lenv}.
Finally, I apply these new environments to the syntax of $\name$ in
\cref{sec:lrsub} to derive renaming and substitution operators.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}

>>>>

The rest of this chapter proceeds as follows.
In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared in my previous work~\citep{WA21}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
The idea of \name{} is to augment the simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
Given this new calculus $\name$, the first goal is to apply the techniques of
\cref{sec:simple} to it, yielding a simultaneous substitution operation.
To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
the typing rules of \name{} to not mention contexts explicitly, as was the style
in \cref{sec:simple}.
This new notation --- the \emph{bunched connectives} --- is versatile at
defining simply typed usage-aware syntaxes, and I give further non-$\name$
examples in \cref{sec:rec}.
Finally, I justify connections to linear logic and modal logic in
\cref{sec:translation}, where I translate $\name$ terms to and from
DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
%The application of the techniques of \cref{sec:simple} is left to
%\cref{sec:ren-sub-lr}.

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{Additions to and variations of \name{}}\label{sec:variant}
\input{rec.tex}
\section{Translation to and from existing systems}\label{sec:translation}
\input{trans.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

@@

In \cref{sec:semirings}, I defined my calculus of interest $\name$.
In this chapter, I develop the necessary syntactic metatheory for
specifying and implementing the substitution operation.
I follow the approach of \cref{sec:kits} using syntactic kits, but have to make
significant changes to the underlying notion of \emph{environment} before doing
so.
I give and informally motivate these changes to environments in
\cref{sec:lrkits}, and prove some properties of the new definition in
\cref{sec:lenv}.
Finally, I apply these new environments to the syntax of $\name$ in
\cref{sec:lrsub} to derive renaming and substitution operators.

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Conclusion}\label{sec:ren-sub-lr-conc}

In the preceding two chapters, I have developed a discipline for specifying
the syntax of linear and modal type systems, and furthermore developing the
syntactic metatheory of those type systems.
All of these are based on semirings, and the linear algebra arising from
considering a usage context full of semiring elements as a vector.

These developments can be seen in retrospect as a generalisation of the methods
explained in \cref{sec:simple}.
In terms of premise connectives in the syntactic rules, we have generalised from
just $\{\dot1, \dottimes\}$ to
$\{\dot1, \dottimes, I^*, \sep, \gr r\cdot{}, \Box^{0{+}}\}$, maintaining our
ability to keep the context implicit.
Similarly to how rule premises can require separation of usage annotations, our
new environments can require such a separation between their entries thanks to
the linear map they now contain.
I have generalised the key property of a kit from arbitrary weakening to
weakening by $\gr0$-annotated variables, and using that have produced a
substitution operation based on the same principles as that from
\cref{sec:kits}.

Having generalised all of the components --- namely the contexts, the syntax,
and the notion of environment --- the type of the substitution operation looks
the same as it did for intuitionistic ST$\lambda$C\@.
Being able to maintain this uniformity is a key step towards generalising the
rest of \cref{sec:simple} (i.e., \cref{sec:gen-sem,sec:gen-syn}), as I do in
\cref{sec:framework}.

A similar substitution lemma appears in the PhD thesis of
\citet[p.\ 138]{petricek-thesis} under the name \emph{multi-nary substitution}.
In my notation, \citeauthor{petricek-thesis}'s substitution rule looks like the
following, up to permutation of the contexts containing $\Gamma$.
Note that if $\Delta = \grQ\delta$, then $\gr r\Delta$ denotes the context
$\plr{\gr r\grQ}\delta$.
This rule is essentially an iterated version of the standard linear single
substitution principle, and is used by \citeauthor{petricek-thesis} as a
strengthened induction hypothesis required to derive single substitution.

\[
  \ebrule{%
    \hypo{\Delta_1 \vdash A_1}
    \hypo{\cdots}
    \hypo{\Delta_n \vdash A_n}
    \hypo{\Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n \vdash B}
    \infer4{\Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n \vdash B}
  }
\]

We can derive \citeauthor{petricek-thesis}-style multi-nary substitution as a
corollary of my simultaneous substitution, using reasoning similar to that of
\cref{thm:single-sub}.

\begin{corollary}\label{thm:petricek-sub}
  \Citeauthor{petricek-thesis}'s multi-nary substitution, as stated above, is
  admissible in $\name$.
\end{corollary}
\begin{proof}
  It is enough to provide a substitution of type
  \[
    \Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n
    \env\vdash \Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n.
  \]
  To do this, we use \cref{thm:construct-env} repeatedly, leaving us needing a
  substitution of type
  $\Gamma, \gr0\Delta_1, \ldots, \gr0\Delta_n \env\vdash \Gamma$ and terms of
  types
  \begin{align*}
    \gr0\gamma, \Delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \gr0\delta_n
    \vdash A_1 \\
    &\vdots \\
    \gr0\gamma, \gr0\delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \Delta_n
    \vdash A_n.
  \end{align*}
  The identity substitution and weakening by $\gr0$-annotated variables is
  enough to make these requirements line up with the given hypotheses.
\end{proof}

My substitution principle is stronger than \citeauthor{petricek-thesis}'s.
Where \citeauthor{petricek-thesis} requires that distinct variables be
available for each hypothesis, I allow for separation of uses via addition of
contexts.
Below is a prototypical example.

\begin{example}
  Let $\Ann \coloneqq \plr{\mathbb N, =, 0, +, 1, \times}$, the exact
  usage-counting posemiring.
  Then, we can construct a substitution $\rho : \gr2A \env\vdash \gr1A, \gr1A$,
  yielding a transformation of terms of the following form:
  \[
    \ebrule{%
      \hypo{\gr1A, \gr1A \vdash B}
      \infer1{\gr2A \vdash B}
    }.
  \]
  To construct $\rho$, we use \cref{thm:construct-env} case
  $\sep(\rightarrowtriangle)$, using the fact that $\gr2 \leq \gr1 + \gr1$.
  From there, two identity substitutions suffice.
  The action of $\rho$ on terms is to merge the two variables into one.
  Note that a renaming, rather than a substitution, would also suffice.
\end{example}

Most notably, my (single) substitution principle more naturally fits the
requirement we would have for the reduct of the $\beta$-rule for functions in
$\name$, whereas \citeauthor{petricek-thesis}'s substitution principle would
need some additional transformation for it to fit properly.
This comes from the fact that the $\name$ function application rule introduces
an algebraic ($+$) separation between its premises, whereas
\citeauthor{petricek-thesis}'s substitution principle separates premises only
via concatenation.
